---
title: "573 Research Project"
author: "Troy Hall"
date: "4/6/2022"
output: html_document
---
### USER NOTE ###
Be sure to set all appropriate filepaths in the chunk below BEFORE running this code. 
Save all dependencies including the 2019 AESC CPS .dat file and .xml file, county data files, "IPUMS health var table" and "1990 Census to NAICS" to working directory BEFORE running this code.

If you focus on the regression prediction problem, please report results for the following
methods: linear regression, LASSO/ridge/elastic net, regression trees and random forests,
boosting, SVM, kNN.

If you focus on the classification problem, please report results for the following meth-
ods: logistic regression, LDA/QDA, classification trees and random forests, boosting, SVM,
and kNN.
```{r}
setwd("C:/Users/graha/Dropbox/PC/Desktop/econ 573/final project")

library(dplyr)
library(tidyverse)
library(ipumsr)
library(readxl)
library(glmnet)
library(ggplot2)
library(pls)
library(class)
library(corrplot)
library(e1071)
library(leaps)
library(tree)
library(caret)
library(textir)
library(GGally)
library(gamlr)
library(ggfortify)
```


# Data import
```{r}
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("cps_00009.xml")
cpsdata <- read_ipums_micro(ddi)
```

# Data Cleaning
```{r}
# Adults
cpsdata = subset(cpsdata, AGE >= 18)

# Industry movement variable
cpsdata$indmove = ifelse(cpsdata$IND1990==cpsdata$IND90LY, 1, 0)

# Live in a rural county (less than 100k people total)
cpsdata$rural = ifelse(cpsdata$COUNTY==0, 1, 0)

# Manufacturing workers
censustonaics = read_xlsx("1990censustoNAICS.xlsx")
censustonaics$`1990 Census` = as.numeric(censustonaics$`1990 Census`)
cpsdata = cpsdata %>%
  rename("1990 Census" = "IND1990")
cpsdata = cpsdata %>% 
  left_join(censustonaics, by = "1990 Census")
manfacNAICS = c(11,21,31,32,33)
cpsdata$manufacturing = ifelse(cpsdata$`1997 NAICS` %in% manfacNAICS, 1, 0)

# Health variable from 1 to 5 to continuous measure between 0 and 1
cpsdata$HEALTH = ifelse(cpsdata$HEALTH == 1, 1,
       ifelse(cpsdata$HEALTH == 2, 0.75,
              ifelse(cpsdata$HEALTH == 3, 0.5,
                     ifelse(cpsdata$HEALTH == 4, 0.25,
                            ifelse(cpsdata$HEALTH == 5, 0, NA)))))

# Reconciling SCHLCOLL var to a single category of not-in-school
cpsdata$SCHLCOLL = ifelse(cpsdata$SCHLCOLL==5, 0, cpsdata$SCHLCOLL)

# Categorizing EDUC into Below HS, HS, Some college, Associates, Bachelors, Graduate training
cpsdata$EDUC = ifelse(cpsdata$EDUC == 91, 92, cpsdata$EDUC)
cpsdata$Education = ifelse(cpsdata$EDUC < 73, 1, 
                               ifelse(cpsdata$EDUC == 73, 2, 
                                      ifelse(cpsdata$EDUC == 81, 3, 
                                             ifelse(cpsdata$EDUC == 92, 4,
                                                    ifelse(cpsdata$EDUC == 111, 5,
                                                           ifelse(cpsdata$EDUC > 111, 6, cpsdata$EDUC))))))

# Adding GDP per capita by state (Probably shouldn't use this variable. Population/GDP numbers are by state due to unavailable counties)
countypopdata = read_excel(path = "County data 2001-19 (Autosaved).xlsx", sheet = 1)
countyGDPdata = read_csv("County GDP data 2001-19.csv")
countypopdata = countypopdata[c("GeoFIPS","GeoName","LineCode","2019")]
countypopdata = countypopdata %>%
  rename("Population2019" = "2019")
countypopdata$GeoFIPS = str_sub(countypopdata$GeoFIPS,2,−2)
countyGDPdata = countyGDPdata[c("GeoFIPS","GeoName","LineCode","2019")]
countyGDPdata = countyGDPdata %>%
  rename("GDP2019" = "2019")

countypopdata = subset(countypopdata, countypopdata$LineCode %in% 10)
countyGDPdata = subset(countyGDPdata, countyGDPdata$LineCode %in% 10)

countydata = merge(countypopdata, countyGDPdata, by.x = "GeoFIPS", by.y = "GeoFIPS")
countydata = countydata[c("GeoFIPS","GeoName.x","LineCode.x","Population2019", "GDP2019")]

cpsdata$COUNTY = ifelse(cpsdata$COUNTY == 00000, cpsdata$STATEFIP, cpsdata$COUNTY)
cpsdata$COUNTY = str_pad(cpsdata$COUNTY, width=5, side="right", pad="0")
cpsdata$STATEFIP = str_pad(cpsdata$STATEFIP, width=5, side="right", pad="0")

countydata$Population2019 = as.numeric(countydata$Population2019)
countydata$GDP2019 = as.numeric(countydata$GDP2019)
countydata = transform(countydata, GDPpercap = GDP2019*1000 / Population2019)

cpsdata1 = merge(cpsdata, countydata, by.x = "STATEFIP", by.y = "GeoFIPS")


# Adding job status vars
cpsdata1$changedjob = ifelse(cpsdata1$OCC1990 == cpsdata1$OCC90LY, 1,
                             ifelse(cpsdata1$OCC1990 != cpsdata1$OCC90LY, 2, 0))

cpsdata1$lostjob1 = ifelse(cpsdata1$OCC1990 == 999, 2, 0)
cpsdata1$lostjob2 = ifelse(cpsdata1$OCC90LY == 999, 1, 0)

cpsdata1$unempoversamp = ifelse(cpsdata1$OCC1990 + cpsdata1$OCC90LY == 3, 1, 0)
cpsdata1$lostjob = ifelse(cpsdata1$OCC90LY - cpsdata1$OCC1990 == -2, 1, 0)

# Race recode
cpsdata1$RACE = ifelse(cpsdata1$RACE == 999, 0, cpsdata1$RACE)
cpsdata1$RACE = ifelse(cpsdata1$RACE == 100, "White", 
                    ifelse(cpsdata1$RACE == 200, "Black",
                        ifelse(cpsdata1$RACE == 651, "Asian",
                            ifelse(cpsdata1$RACE == 652, "Asian", "Other"))))

# Hispanic
cpsdata1$HISPAN = ifelse(cpsdata1$HISPAN > 0, 1, 0)

# Married or not
cpsdata1$MARST = ifelse(cpsdata1$MARST > 2, 1, 0)

# Population status
cpsdata1$POPSTAT = ifelse(cpsdata1$POPSTAT == 1, "Civilian",
                         ifelse(cpsdata1$POPSTAT == 2, "Military",
                                ifelse(cpsdata1$POPSTAT == 3, "Child", 0)))

# Citizen
cpsdata1$CITIZEN = ifelse(cpsdata1$CITIZEN == 5, 0, 1)

# Nativity
cpsdata1$NATIVITY = ifelse(cpsdata1$NATIVITY == 0, 1,
                          ifelse(cpsdata1$NATIVITY == 5, 1, 0))
cpsdata1 = cpsdata1 %>%
  rename("bornabroad" = "NATIVITY")

# Empstat
cpsdata1$EMPSTAT = ifelse(cpsdata1$EMPSTAT == 1, "Military",
                         ifelse(cpsdata1$EMPSTAT == 10, "Employed",
                                ifelse(cpsdata1$EMPSTAT == 12, "Employed",
                                       ifelse(cpsdata1$EMPSTAT == 21, "Unemployed",
                                              ifelse(cpsdata1$EMPSTAT == 22, "Unemployed",
                                                     ifelse(cpsdata1$EMPSTAT > 29, "NILF", NA))))))

# Labor force or NILF
cpsdata1$LABFORCE = ifelse(cpsdata1$LABFORCE == 2, 1, 0)

# Migration recode
cpsdata1$MIGRATE1 = ifelse(cpsdata1$MIGRATE1 == 1, "Same residence",
       ifelse(cpsdata1$MIGRATE1 == 3, "Moved, same county",
              ifelse(cpsdata1$MIGRATE1 == 4, "Moved, diff county",
                     ifelse(cpsdata1$MIGRATE1 == 5, "Moved, new state",
                            ifelse(cpsdata1$MIGRATE1 == 6, "Moved, abroad", 0)))))

# Disability recode
cpsdata1$DISABWRK = ifelse(cpsdata1$DISABWRK == 2, 1, 0)

# Adjusting income vars to non-negative and logging
cpsdata1$HHINCOME = log(cpsdata1$HHINCOME + abs(min(cpsdata1$HHINCOME)) + 1)
cpsdata1$FTOTVAL = log(cpsdata1$FTOTVAL + abs(min(cpsdata1$FTOTVAL)) + 1)
cpsdata1$INCTOT = log(cpsdata1$INCTOT + abs(min(cpsdata1$INCTOT)) + 1)



# Keeping useful vars/cleanup
cpsdata1 = cpsdata1[c("HEALTH", "RACE", "HISPAN", "SEX", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "PERNUM", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]

# Cleaning up hours worked
cpsdata1$UHRSWORKT = ifelse(cpsdata1$UHRSWORKT > 169, 0, cpsdata$UHRSWORKT)

# Metropolitan area (1 for metro area, 0 for not)
cpsdata1$METRO = ifelse(cpsdata1$METRO == 0, 0, 
                        ifelse(cpsdata1$METRO == 1, 0, 
                               ifelse(cpsdata1$METRO > 1.5, 1, 0)))

# Binary sex
cpsdata1$SEX = ifelse(cpsdata1$SEX == 1, 0,
                      ifelse(cpsdata1$SEX == 2, 1, 0))
cpsdata1 = cpsdata1 %>%
  rename("FEMALE" = "SEX")

# Binary health var
cpsdata1$goodhealth = ifelse(cpsdata1$HEALTH > 0.26, 1, 0)

cpsdataoriginaldataset = cpsdata
cpsdata = cpsdata1
cpsdata1
```

# Setting X vars and Y of health
```{r}
cpshealth = cpsdata1[cpsdata1$HEALTH>0, ]
cpshealth = cpsdata1[cpsdata1$AGE>0, ]
cpshealth = cpsdata1[cpsdata1$FAMSIZE>0, ]
cpshealth = cpsdata1[cpsdata1$Education>1, ]
cpshealth = cpsdata1[cpsdata1$HHINCOME>=0, ]

cpsmetro = cpshealth[cpsdata1$METRO ==1, ]

metroasian = cpsmetro[cpsmetro$RACE =="Other", ]

x = data.matrix(cpshealth[ c("RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")])
y = cpshealth$HEALTH

dfhealth = cpshealth %>%
  select(RACE, HISPAN, FEMALE, MARST, METRO, Education, STATEFIP, HHINCOME, AGE, VETSTAT, FAMSIZE, LABFORCE, bornabroad, UHRSWORKT, indmove, lostjob, ANYCOVNW, HEALTH )

dfmetro = cpsmetro %>%
  select(RACE, HISPAN, FEMALE, MARST, METRO, Education, STATEFIP, HHINCOME, AGE, VETSTAT, FAMSIZE, LABFORCE, bornabroad, UHRSWORKT, indmove, lostjob, ANYCOVNW, HEALTH )

dfhealth_numeric = cpshealth %>%
  select(HISPAN, FEMALE, MARST, METRO, Education, HHINCOME, AGE, VETSTAT, FAMSIZE, LABFORCE, bornabroad, UHRSWORKT, indmove, lostjob, ANYCOVNW, HEALTH )

dfmetro = cpsmetro %>%
  select(RACE, HISPAN, FEMALE, MARST, METRO, Education, STATEFIP, HHINCOME, AGE, VETSTAT, FAMSIZE, LABFORCE, bornabroad, UHRSWORKT, indmove, lostjob, ANYCOVNW, HEALTH )
```


##############################################################################

End cleaning, begin models

##############################################################################



#summary stats
```{r}

colMeans(dfhealth[sapply(dfhealth, is.numeric)])

colMeans(dfmetro[sapply(dfmetro, is.numeric)]) 


```


# OLS
```{r}

OLS = lm(y ~ x, data = cpshealth)
OLS
```

# Ridge
```{r}
ridge.model = cv.glmnet(x,y, type.measure='mse', alpha=0)

ridge.model$lambda.min
#reports lowest at 0.0083

#plot it
plot(ridge.model)

#selects 8 model but with log-L of -3.something
#coef(ridge.model)

new_ridge = glmnet(x,y,alpha=0, lambda =ridge.model$lambda.min )
coef(new_ridge)
```
# LASSO
```{r}
lasso.model = cv.glmnet(x, y, type.measure = "mse", alpha=1)

coef(lasso.model)


lowest_lambda_lasso = lasso.model$lambda.min
lowest_lambda_lasso
#0.0002177337

lasso.model_withlowestlambda = glmnet(x,y, alpha=1, lambda = lowest_lambda_lasso)

#checking coef's for lasso
coef(lasso.model_withlowestlambda)



#plot it
plot(lasso.model)

#predicts lowest mse with 9 and 16 variable model

```



# PCR not needed but interesting
```{r}
pca.fit <- prcomp(dfhealth_numeric, center = TRUE,scale. = TRUE)
pca.fit
autoplot(pca.fit, loadings = TRUE)

```


# elastic net
```{r}
#create train and test data

trainIndex_health <- createDataPartition(dfhealth$HEALTH, p = .8,
                                  list = FALSE,
                                  times = 1)
vector_train_index =c(trainIndex_health)
train_elastic <- dfhealth[ vector_train_index,]
valid_elastic <- dfhealth[-vector_train_index,]

#creating model

elastic.model = train(
  HEALTH~., data = train_elastic, method = "glmnet",
  trControl = trainControl("cv", number = 10),
  tuneLength = 10
)
elastic.model$bestTune
plot(elastic.model)

#reports an optimal alpha of 0.8 and lambda of 0.0004774163

coef(elastic.model$finalModel, elastic.model$bestTune$lambda)


```


# KNN
#```{r}
# Prepping the dataset (no factor vars)
#cpsKNNt = cpshealth[complete.cases(cpshealth), ]
#cpsKNN = cpshealth[c("RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]
#cpsKNN$RACE = as.factor(cpsKNN$RACE)
#cpsKNN$RACE = as.numeric(cpsKNN$RACE)
#cpsKNN$AGE = as.numeric(cpsKNN$AGE)

# Prepping KNN
#samp = sample(1:nrow(cpsKNN), 0.5 * nrow(cpsKNN))
#knn.train = cpsKNN[samp,]
#knn.test = cpsKNN[-samp,]
#train.health = cpsKNNt$HEALTH[samp]
#test.health = cpsKNNt$HEALTH[-samp]

# Running KNN and reporting results
#knn.results = data.frame(k = 1:10, testerror = NA)
#for(i in 1:10){
#knn.pred <- knn(knn.train, knn.test, train.health, k = i)
#knn.results$testerror[i] = mean(knn.pred == test.health)}

#knn.results
#```




# SVM
```{r}
#copy for experimenting
df_svm_full = dfhealth
set.seed(123)

#svm takes far too long for a laptop to compute 120,000 observations so will try half of that number in a random sample
#linear:50000
#radial: 20000
#poly: less than 5k
df_svm = sample_n(df_svm_full, 20000)

#changing y to factor variable
df_svm$HEALTH <- as.factor(df_svm$HEALTH)

# ################### creating training data
intrain <- createDataPartition(y= df_svm$HEALTH, p = .8,
                                  list = FALSE,
                                  times = 1)
svm_train_index =c(intrain)
svm_training <- df_svm[svm_train_index, ]
svm_testing <- df_svm[-svm_train_index, ]

#checking dimensions
#dim(svm_training)
#dim(svm_testing)

#convert y into factor
svm_training[["HEALTH"]] = factor(svm_training[["HEALTH"]])


# ####### ############# ################################  LINEAR


svm_linear = train(
  HEALTH~., data = svm_training, method = "svmLinear", 
  trControl = trainControl("cv", number = 10), 
  tuneGrid = expand.grid(C = seq(0, 2, length = 20)), 
  preProcess = c("center", "scale"), tuneLength = 10
  )
svm_linear
plot(svm_linear)

test_pred <- predict(svm_linear, newdata = svm_testing)
test_pred

confusionMatrix(table(test_pred, svm_testing$HEALTH))
svm_linear$bestTune
 
mean(test_pred == svm_testing$HEALTH)


############################################################ RADIAL
 
svm_radial = train(
  HEALTH~., data = svm_training, method = "svmRadial", 
  trControl = trainControl("cv", number = 10), 
  preProcess = c("center","scale"), 
  tuneLength = 10
  )

svm_radial

svm_radial$bestTune
test_pred_radial <- predict(svm_radial, newdata = svm_testing)
test_pred_radial

confusionMatrix(table(test_pred_radial, svm_testing$HEALTH))
mean(test_pred_radial == svm_testing$HEALTH)

##############################################################  POLY
svm_poly = train(
  HEALTH~., data = svm_training, method = "svmPoly", 
  trControl = trainControl("cv", number = 10), 
  preProcess = c("center","scale"), 
  tuneLength = 10
  )

svm_poly

svm_poly$bestTune
test_pred_poly <- predict(svm_poly, newdata = svm_testing)
test_pred_poly

confusionMatrix(table(test_pred_poly, svm_testing$HEALTH))
mean(test_pred_poly == svm_testing$HEALTH)


# ######################################


```


