---
title: "573 Research Project"
author: "Troy Hall"
date: "4/6/2022"
output: html_document
---
### USER NOTE ###
Be sure to set all appropriate filepaths in the chunk below BEFORE running this code. 
Save all dependencies including the 2019 AESC CPS .dat file and .xml file, county data files, "IPUMS health var table" and "1990 Census to NAICS" to working directory BEFORE running this code.

If you focus on the regression prediction problem, please report results for the following
methods: linear regression, LASSO/ridge/elastic net, regression trees and random forests,
boosting, SVM, kNN.

If you focus on the classification problem, please report results for the following meth-
ods: logistic regression, LDA/QDA, classification trees and random forests, boosting, SVM,
and kNN.
```{r}
workingd = c("C:/Users/troyhall/Documents/Classes/Spring '22/ECON 573/Final Project/")
setwd(workingd)
library(dplyr)
library(tidyverse)
library(ipumsr)
library(readxl)
library(glmnet)
library(ggplot2)
library(pls)
library(class)
library(corrplot)
library(e1071)
library(leaps)
library(tree)
```


# Data import
```{r}
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("cps_00009.xml")
cpsdata <- read_ipums_micro(ddi)
```

# Data Cleaning
```{r}
# Adults
cpsdata = subset(cpsdata, AGE >= 18)

# Industry movement variable
cpsdata$indmove = ifelse(cpsdata$IND1990==cpsdata$IND90LY, 1, 0)

# Live in a rural county (less than 100k people total)
cpsdata$rural = ifelse(cpsdata$COUNTY==0, 1, 0)

# Manufacturing workers
censustonaics = read_xlsx("1990censustoNAICS.xlsx")
censustonaics$`1990 Census` = as.numeric(censustonaics$`1990 Census`)
cpsdata = cpsdata %>%
  rename("1990 Census" = "IND1990")
cpsdata = cpsdata %>% 
  left_join(censustonaics, by = "1990 Census")
manfacNAICS = c(11,21,31,32,33)
cpsdata$manufacturing = ifelse(cpsdata$`1997 NAICS` %in% manfacNAICS, 1, 0)

# Health variable from 1 to 5 to continuous measure between 0 and 1
cpsdata$HEALTH = ifelse(cpsdata$HEALTH == 1, 1,
       ifelse(cpsdata$HEALTH == 2, 0.75,
              ifelse(cpsdata$HEALTH == 3, 0.5,
                     ifelse(cpsdata$HEALTH == 4, 0.25,
                            ifelse(cpsdata$HEALTH == 5, 0, NA)))))

# Reconciling SCHLCOLL var to a single category of not-in-school
cpsdata$SCHLCOLL = ifelse(cpsdata$SCHLCOLL==5, 0, cpsdata$SCHLCOLL)

# Categorizing EDUC into Below HS, HS, Some college, Associates, Bachelors, Graduate training
cpsdata$EDUC = ifelse(cpsdata$EDUC == 91, 92, cpsdata$EDUC)
cpsdata$Education = ifelse(cpsdata$EDUC < 73, 1, 
                               ifelse(cpsdata$EDUC == 73, 2, 
                                      ifelse(cpsdata$EDUC == 81, 3, 
                                             ifelse(cpsdata$EDUC == 92, 4,
                                                    ifelse(cpsdata$EDUC == 111, 5,
                                                           ifelse(cpsdata$EDUC > 111, 6, cpsdata$EDUC))))))

# Adding GDP per capita by state (Probably shouldn't use this variable. Population/GDP numbers are by state due to unavailable counties)
countypopdata = read_excel(path = "County data 2001-19 (Autosaved).xlsx", sheet = 1)
countyGDPdata = read_csv("County GDP data 2001-19.csv")
countypopdata = countypopdata[c("GeoFIPS","GeoName","LineCode","2019")]
countypopdata = countypopdata %>%
  rename("Population2019" = "2019")
countypopdata$GeoFIPS = str_sub(countypopdata$GeoFIPS,2,−2)
countyGDPdata = countyGDPdata[c("GeoFIPS","GeoName","LineCode","2019")]
countyGDPdata = countyGDPdata %>%
  rename("GDP2019" = "2019")

countypopdata = subset(countypopdata, countypopdata$LineCode %in% 10)
countyGDPdata = subset(countyGDPdata, countyGDPdata$LineCode %in% 10)

countydata = merge(countypopdata, countyGDPdata, by.x = "GeoFIPS", by.y = "GeoFIPS")
countydata = countydata[c("GeoFIPS","GeoName.x","LineCode.x","Population2019", "GDP2019")]

cpsdata$COUNTY = ifelse(cpsdata$COUNTY == 00000, cpsdata$STATEFIP, cpsdata$COUNTY)
cpsdata$COUNTY = str_pad(cpsdata$COUNTY, width=5, side="right", pad="0")
cpsdata$STATEFIP = str_pad(cpsdata$STATEFIP, width=5, side="right", pad="0")

countydata$Population2019 = as.numeric(countydata$Population2019)
countydata$GDP2019 = as.numeric(countydata$GDP2019)
countydata = transform(countydata, GDPpercap = GDP2019*1000 / Population2019)

cpsdata1 = merge(cpsdata, countydata, by.x = "STATEFIP", by.y = "GeoFIPS")


# Adding job status vars
cpsdata1$changedjob = ifelse(cpsdata1$OCC1990 == cpsdata1$OCC90LY, 1,
                             ifelse(cpsdata1$OCC1990 != cpsdata1$OCC90LY, 2, 0))

cpsdata1$lostjob1 = ifelse(cpsdata1$OCC1990 == 999, 2, 0)
cpsdata1$lostjob2 = ifelse(cpsdata1$OCC90LY == 999, 1, 0)

cpsdata1$unempoversamp = ifelse(cpsdata1$OCC1990 + cpsdata1$OCC90LY == 3, 1, 0)
cpsdata1$lostjob = ifelse(cpsdata1$OCC90LY - cpsdata1$OCC1990 == -2, 1, 0)

# Race recode
cpsdata1$RACE = ifelse(cpsdata1$RACE == 999, 0, cpsdata1$RACE)
cpsdata1$RACE = ifelse(cpsdata1$RACE == 100, "White", 
                    ifelse(cpsdata1$RACE == 200, "Black",
                        ifelse(cpsdata1$RACE == 651, "Asian",
                            ifelse(cpsdata1$RACE == 652, "Asian", "Other"))))

# Hispanic
cpsdata1$HISPAN = ifelse(cpsdata1$HISPAN > 0, 1, 0)

# Married or not
cpsdata1$MARST = ifelse(cpsdata1$MARST > 2, 1, 0)

# Population status
cpsdata1$POPSTAT = ifelse(cpsdata1$POPSTAT == 1, "Civilian",
                         ifelse(cpsdata1$POPSTAT == 2, "Military",
                                ifelse(cpsdata1$POPSTAT == 3, "Child", 0)))

# Citizen
cpsdata1$CITIZEN = ifelse(cpsdata1$CITIZEN == 5, 0, 1)

# Nativity
cpsdata1$NATIVITY = ifelse(cpsdata1$NATIVITY == 0, 1,
                          ifelse(cpsdata1$NATIVITY == 5, 1, 0))
cpsdata1 = cpsdata1 %>%
  rename("bornabroad" = "NATIVITY")

# Empstat
cpsdata1$EMPSTAT = ifelse(cpsdata1$EMPSTAT == 1, "Military",
                         ifelse(cpsdata1$EMPSTAT == 10, "Employed",
                                ifelse(cpsdata1$EMPSTAT == 12, "Employed",
                                       ifelse(cpsdata1$EMPSTAT == 21, "Unemployed",
                                              ifelse(cpsdata1$EMPSTAT == 22, "Unemployed",
                                                     ifelse(cpsdata1$EMPSTAT > 29, "NILF", NA))))))

# Labor force or NILF
cpsdata1$LABFORCE = ifelse(cpsdata1$LABFORCE == 2, 1, 0)

# Migration recode
cpsdata1$MIGRATE1 = ifelse(cpsdata1$MIGRATE1 == 1, "Same residence",
       ifelse(cpsdata1$MIGRATE1 == 3, "Moved, same county",
              ifelse(cpsdata1$MIGRATE1 == 4, "Moved, diff county",
                     ifelse(cpsdata1$MIGRATE1 == 5, "Moved, new state",
                            ifelse(cpsdata1$MIGRATE1 == 6, "Moved, abroad", 0)))))

# Disability recode
cpsdata1$DISABWRK = ifelse(cpsdata1$DISABWRK == 2, 1, 0)

# Adjusting income vars to non-negative and logging
cpsdata1$HHINCOME = log(cpsdata1$HHINCOME + abs(min(cpsdata1$HHINCOME)) + 1)
cpsdata1$FTOTVAL = log(cpsdata1$FTOTVAL + abs(min(cpsdata1$FTOTVAL)) + 1)
cpsdata1$INCTOT = log(cpsdata1$INCTOT + abs(min(cpsdata1$INCTOT)) + 1)



# Keeping useful vars/cleanup
cpsdata1 = cpsdata1[c("HEALTH", "RACE", "HISPAN", "SEX", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "PERNUM", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]

# Cleaning up hours worked
cpsdata1$UHRSWORKT = ifelse(cpsdata1$UHRSWORKT > 169, 0, cpsdata$UHRSWORKT)

# Metropolitan area (1 for metro area, 0 for not)
cpsdata1$METRO = ifelse(cpsdata1$METRO == 0, 0, 
                        ifelse(cpsdata1$METRO == 1, 0, 
                               ifelse(cpsdata1$METRO > 1.5, 1, 0)))

# Binary sex
cpsdata1$SEX = ifelse(cpsdata1$SEX == 1, 0,
                      ifelse(cpsdata1$SEX == 2, 1, 0))
cpsdata1 = cpsdata1 %>%
  rename("FEMALE" = "SEX")

# Binary health var
cpsdata1$goodhealth = ifelse(cpsdata1$HEALTH > 0.26, 1, 0)

cpsdataoriginaldataset = cpsdata
cpsdata = cpsdata1
cpsdata1
```

# Setting X vars and Y of health
```{r}
cpshealth = cpsdata1[cpsdata1$HEALTH>0, ]
cpshealth = cpsdata1[cpsdata1$AGE>0, ]
cpshealth = cpsdata1[cpsdata1$FAMSIZE>0, ]
cpshealth = cpsdata1[cpsdata1$Education>1, ]
cpshealth = cpsdata1[cpsdata1$HHINCOME>=0, ]
x = data.matrix(cpshealth[ c("RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")])
y = cpshealth$HEALTH
```


##############################################################################

End cleaning, begin models

##############################################################################


# OLS
```{r}
OLS = lm(y ~ x, data = cpshealth)
```

# LASSO
```{r}
lasso.model = cv.glmnet(x, y, type.measure = "mse")

#plot it
plot(lasso.model)
#predicts lowest mse with 5 variable model

#checking coef's for lasso
coef(lasso.model)
```

# Ridge
```{r}
ridge.model = cv.glmnet(x,y, type.measure='mse', alpha=0)
#plot it
plot(ridge.model)
#selects 8 model but with log-L of -3.something
coef(ridge.model)
```

# PCR
```{r}
pcr.fit <- pcr(y ~ x, data = cpshealth, scale = TRUE, validation = "CV")
```

# KNN
```{r}
# Prepping the dataset (no factor vars)
cpsKNNt = cpshealth[complete.cases(cpshealth), ]
cpsKNN = cpshealth[c("RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]
cpsKNN$RACE = as.factor(cpsKNN$RACE)
cpsKNN$RACE = as.numeric(cpsKNN$RACE)
cpsKNN$AGE = as.numeric(cpsKNN$AGE)

# Prepping KNN
samp = sample(1:nrow(cpsKNN), 0.5 * nrow(cpsKNN))
knn.train = cpsKNN[samp,]
knn.test = cpsKNN[-samp,]
train.health = cpsKNNt$HEALTH[samp]
test.health = cpsKNNt$HEALTH[-samp]

# Running KNN and reporting results
knn.results = data.frame(k = 1:10, testerror = NA)
for(i in 1:10){
knn.pred <- knn(knn.train, knn.test, train.health, k = i)
knn.results$testerror[i] = mean(knn.pred == test.health)
}
```

# Trees
```{r}
cpsdatatrees = cpsdata1[c("HEALTH", "RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]
cpsdatatrees$HEALTH <-factor(cpsdatatrees$HEALTH)

#create a training set for the trees
traintrees <- sample (1: nrow (cpsdatatrees), nrow (cpsdatatrees) / 2)
#create a decision tree no subset selection
fulldatasettree <- tree(HEALTH~., cpsdatatrees, subset=traintrees)
#summary and plot the decision tree
summary(fulldatasettree)
plot(fulldatasettree)
text(fulldatasettree, pretty=0)
#create another decision tree by using cross validation
cv.tree <- cv.tree(fulldatasettree)
plot (cv.tree$size, cv.tree$dev, type = "b")

tmin <- which.min(cv.tree$dev)
tmin

#prune the tree using the cross validtion
prunedtree <- prune.tree (fulldatasettree , best = tmin)
plot (prunedtree)
text (prunedtree , pretty = 0)

#predicted vlaues for the test data using full tree
yhat <- predict (fulldatasettree , newdata = cpsdatatrees[-traintrees ,])
#predicted vlaues for the test data using prunded tree
yhat2 <- predict (prunedtree , newdata = cpsdatatrees[-traintrees ,])

#plot actual vs predicted and find MSE
testtrees <- cpsdatatrees[-traintrees, "HEALTH"]
plot (yhat , testtrees)

abline (0, 1)
mean ((yhat - testtrees)^2)

plot (yhat2 , testtrees)

abline (0, 1)
mean ((yhat2 - testtrees)^2)
```



#Bagging 
```{r}
cpsdatabag = cpsdata1[c("HEALTH", "RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]
cpsdatabag$HEALTH <-factor(cpsdatabag$HEALTH)
library(randomForest)
trainbag <- sample (1: nrow (cpsdatabag), nrow (cpsdatabag) / 2)
#Create a random forest using bagging (try all 36 variables)
bag.cps <- randomForest(HEALTH ~., data=cpsdatabag, subset = trainbag, mtry=18, importance = TRUE)

#predict y using bag.cps
yhatbag <- predict (bag.cps, newdata = cpsdatabag[-trainbag ,])

#plot predicted vs acutal for test data
testbag <- cpsdatabag[-trainbag, "HEALTH"]
plot (yhatbag , testbag)

abline (0, 1)
mean((yhatbag - testbag)^2)
```



#Forest
```{r}
cpsdataforest = cpsdata1[c("HEALTH", "RACE", "HISPAN", "FEMALE", "MARST", "METRO", "Education", "STATEFIP", "HHINCOME", "AGE", "VETSTAT", "FAMSIZE", "LABFORCE", "bornabroad", "UHRSWORKT", "indmove", "lostjob", "ANYCOVNW")]
cpsdataforest$HEALTH <-factor(cpsdataforest$HEALTH)
library(randomForest)
trainforest <- sample (1: nrow (cpsdataforest), nrow (cpsdataforest) / 2)
#Create Random Forest
forest.cps <- randomForest(HEALTH ~., data=cpsdataforest, subset = trainforest, mtry=6, importance = TRUE)

yhatforest <- predict (bag.cps, newdata = cpsdataforest[-trainforest ,])

testforest <- cpsdataforest[-trainforest, "HEALTH"]
plot (yhatforest , testforest)

abline (0, 1)
mean ((yhatforest - testforest)^2)
```

# SVM
```{r}
dat = data.frame(x=x, y=as.factor(y))

svm.fit = svm(y~ ., data = dat, kernal = 'linear', cost = 10, scale = TRUE, na.action = na.omit)
```


