---
title: "573 Research Project"
author: "Graham Patton"
date: "4/6/2022"
output: html_document
---
### USER NOTE ###
Be sure to set all appropriate filepaths in the chunk below BEFORE running this code. 
Save all dependencies including the 2019 AESC CPS .dat file and .xml file, county data files, "IPUMS health var table" and "1990 Census to NAICS" to working directory BEFORE running this code.

```{r}
setwd("C:/Users/graha/Dropbox/PC/Desktop/econ 573/final project")

library(dplyr)
library(tidyverse)
library(ipumsr)
library(readxl)
library(glmnet)
library(ggplot2)
library(pls)
library(class)
library(corrplot)
library(e1071)
library(leaps)
```


# Data import
```{r}
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("cps_00007.xml")
cpsdata <- read_ipums_micro(ddi)
```

# Data Cleaning
```{r}
# Adults
cpsdata = subset(cpsdata, AGE >= 18)

# Industry movement variable
cpsdata$indmove = ifelse(cpsdata$IND1990==cpsdata$IND90LY, 1, 0)

# Live in a rural county (less than 100k people total)
cpsdata$rural = ifelse(cpsdata$COUNTY==0, 1, 0)

# Manufacturing workers
censustonaics = read_xlsx("1990censustoNAICS.xlsx")
censustonaics$`1990 Census` = as.numeric(censustonaics$`1990 Census`)
cpsdata = cpsdata %>%
  rename("1990 Census" = "IND1990")
cpsdata = cpsdata %>% 
  left_join(censustonaics, by = "1990 Census")
manfacNAICS = c(11,21,31,32,33)
cpsdata$manufacturing = ifelse(cpsdata$`1997 NAICS` %in% manfacNAICS, 1, 0)

# Health variable from 1 to 5 to continuous measure between 0 and 1
cpsdata$HEALTH = ifelse(cpsdata$HEALTH == 1, 1,
       ifelse(cpsdata$HEALTH == 2, 0.75,
              ifelse(cpsdata$HEALTH == 3, 0.5,
                     ifelse(cpsdata$HEALTH == 4, 0.25,
                            ifelse(cpsdata$HEALTH == 5, 0, NA)))))

# Reconciling SCHLCOLL var to a single category of not-in-school
cpsdata$SCHLCOLL = ifelse(cpsdata$SCHLCOLL==5, 0, cpsdata$SCHLCOLL)

# Categorizing EDUC into Below HS, HS, Some college, Associates, Bachelors, Graduate training
cpsdata$EDUC = ifelse(cpsdata$EDUC == 91, 92, cpsdata$EDUC)
cpsdata$Education = ifelse(cpsdata$EDUC < 73, 1, 
                               ifelse(cpsdata$EDUC == 73, 2, 
                                      ifelse(cpsdata$EDUC == 81, 3, 
                                             ifelse(cpsdata$EDUC == 92, 4,
                                                    ifelse(cpsdata$EDUC == 111, 5,
                                                           ifelse(cpsdata$EDUC > 111, 6, cpsdata$EDUC))))))

# Adding GDP per capita by state (Probably shouldn't use this variable. Population/GDP numbers are by state due to unavailable counties)
countypopdata = read_excel(path = "County data 2001-19 (Autosaved).xlsx", sheet = 1)
countyGDPdata = read_csv("County GDP data 2001-19.csv")
countypopdata = countypopdata[c("GeoFIPS","GeoName","LineCode","2019")]
countypopdata = countypopdata %>%
  rename("Population2019" = "2019")
countypopdata$GeoFIPS = str_sub(countypopdata$GeoFIPS,2,−2)
countyGDPdata = countyGDPdata[c("GeoFIPS","GeoName","LineCode","2019")]
countyGDPdata = countyGDPdata %>%
  rename("GDP2019" = "2019")

countypopdata = subset(countypopdata, countypopdata$LineCode %in% 10)
countyGDPdata = subset(countyGDPdata, countyGDPdata$LineCode %in% 10)

countydata = merge(countypopdata, countyGDPdata, by.x = "GeoFIPS", by.y = "GeoFIPS")
countydata = countydata[c("GeoFIPS","GeoName.x","LineCode.x","Population2019", "GDP2019")]

cpsdata$COUNTY = ifelse(cpsdata$COUNTY == 00000, cpsdata$STATEFIP, cpsdata$COUNTY)
cpsdata$COUNTY = str_pad(cpsdata$COUNTY, width=5, side="right", pad="0")
cpsdata$STATEFIP = str_pad(cpsdata$STATEFIP, width=5, side="right", pad="0")

countydata$Population2019 = as.numeric(countydata$Population2019)
countydata$GDP2019 = as.numeric(countydata$GDP2019)
countydata = transform(countydata, GDPpercap = GDP2019*1000 / Population2019)

cpsdata1 = merge(cpsdata, countydata, by.x = "STATEFIP", by.y = "GeoFIPS")


# Adding job status vars
cpsdata1$changedjob = ifelse(cpsdata1$OCC1990 == cpsdata1$OCC90LY, 1,
                             ifelse(cpsdata1$OCC1990 != cpsdata1$OCC90LY, 2, 0))

cpsdata1$lostjob1 = ifelse(cpsdata1$OCC1990 == 999, 2, 0)
cpsdata1$lostjob2 = ifelse(cpsdata1$OCC90LY == 999, 1, 0)

cpsdata1$unempoversamp = ifelse(cpsdata1$OCC1990 + cpsdata1$OCC90LY == 3, 1, 0)
cpsdata1$lostjob = ifelse(cpsdata1$OCC90LY - cpsdata1$OCC1990 == -2, 1, 0)

# Race recode
cpsdata1$RACE = ifelse(cpsdata1$RACE == 999, 0, cpsdata1$RACE)
cpsdata1$RACE = ifelse(cpsdata1$RACE == 100, "White", 
                               ifelse(cpsdata1$RACE == 200, "Black", 
                                      ifelse(cpsdata1$RACE == 300, "Native", 
                                             ifelse(cpsdata1$RACE == 651, "Asian",
                                                    ifelse(cpsdata1$RACE == 652, "Pacific Islander",
                                                           ifelse(cpsdata1$RACE > 653, "Multiracial", ifelse(cpsdata1$HISPAN != 0, "Hispanic", "Other")))))))

# Married or not
cpsdata1$MARST = ifelse(cpsdata1$MARST > 2, 1, 0)

# Population status

#  R doesn't like this below for popstat. throwing errror below:
#Error in ifelse(cpsdata1$POPSTAT == 3, "Child") :   argument "no" is missing, with no default

cpsdata1$POPSTAT = ifelse(cpsdata1$POPSTAT == 1, "Civilian",
                         ifelse(cpsdata1$POPSTAT == 2, "Military",
                                ifelse(cpsdata1$POPSTAT == 3, "Child")))

# Citizen
cpsdata1$CITIZEN = ifelse(cpsdata1$CITIZEN == 5, 0, 1)

# Nativity
cpsdata1$NATIVITY = ifelse(cpsdata1$NATIVITY == 0, "Unknown",
                          ifelse(cpsdata1$NATIVITY == 5, "Foreign-born", "Native"))

# Empstat
cpsdata1$EMPSTAT = ifelse(cpsdata1$EMPSTAT == 1, "Military",
                         ifelse(cpsdata1$EMPSTAT == 10, "Employed",
                                ifelse(cpsdata1$EMPSTAT == 12, "Employed",
                                       ifelse(cpsdata1$EMPSTAT == 21, "Unemployed",
                                              ifelse(cpsdata1$EMPSTAT == 22, "Unemployed",
                                                     ifelse(cpsdata1$EMPSTAT > 29, "NILF", NA))))))

# Labor force or NILF
cpsdata1$LABFORCE = ifelse(cpsdata1$LABFORCE == 2, 1, 0)

# Migration recode
cpsdata1$MIGRATE1 = ifelse(cpsdata1$MIGRATE1 == 1, "Same residence",
       ifelse(cpsdata1$MIGRATE1 == 3, "Moved, same county",
              ifelse(cpsdata1$MIGRATE1 == 4, "Moved, diff county",
                     ifelse(cpsdata1$MIGRATE1 == 5, "Moved, new state",
                            ifelse(cpsdata1$MIGRATE1 == 6, "Moved, abroad", 0)))))

# Disability recode
cpsdata1$DISABWRK = ifelse(cpsdata1$DISABWRK == 2, 1, 0)

# Keeping useful vars/cleanup
cpsdata1 = cpsdata1[c("STATEFIP","REGION","HHINCOME","NFAMS", "PERNUM", "RELATE", "AGE", "SEX", "RACE", "MARST", "POPSTAT", "VETSTAT", "FAMSIZE", "NCHILD", "NCHLT5", "YRIMMIG", "NATIVITY", "LABFORCE", "UHRSWORKT", "SCHLCOLL", "DIFFANY", "FTOTVAL", "INCTOT", "MIGRATE1", "DISABWRK", "HEALTH", "GOTWIC", "indmove", "rural", "manufacturing", "Education", "Population2019", "GDP2019", "changedjob", "unempoversamp", "lostjob")]
cpsdataoriginaldataset = cpsdata
cpsdata = cpsdata1

```
#checking labels
labels(cpsdata1)



#cv_model = cv.glmnet(x_predictors, y_health, alpha=1)

cpshealth = cpsdata1[cpsdata1$HEALTH>0, ]
cpshealth = cpsdata1[cpsdata1$RACE>50, ]
cpshealth = cpsdata1[cpsdata1$AGE>0, ]
cpshealth = cpsdata1[cpsdata1$FAMSIZE>0, ]
cpshealth = cpsdata1[cpsdata1$NCHILD>=0, ]
cpshealth = cpsdata1[cpsdata1$EDUC>1, ]
cpshealth = cpsdata1[cpsdata1$rural>=0, ]
cpshealth = cpsdata1[cpsdata1$HHINCOME>=0, ]

#create log of income variable
attach(cpshealth)
cpshealth$logHHINCOME = log(cpshealth$HHINCOME)


#check labels
names(cpshealth)

#matrix of x variables to test models code
x = data.matrix(cpshealth[ c('logHHINCOME', 'SCHLCOLL', 'NFAMS', 'SEX', 'AGE', 'FAMSIZE', 'RACE', 'DISABWRK', 'rural')])


#y variables
y = cpshealth$HEALTH

# ######################## this is just a matrix of possible variables we could use. will be finalized later
 
#x = data.matrix(cpshealth[, c("STATEFIP", "SERIAL", "REGION", "COUNTY", "NFAMS", "AGE", "SEX", "RACE", "MARST", "FAMSIZE",
#"NCHILD","CITIZEN", "EMPSTAT", "EDUC", "SCHLCOLL", "DIFFANY", 

#"OCC90LY", "IND90LY","LABFORCE","OCC1990", "1990 Census",

#"WORKLY", "MIGSTA1", "WHYMOVE", "MIGRATE1", "DISABWRK", "indmove", "rural", "manufacturing", "Education_1", 
#"GeoName.x", "LineCode.x", "Population2019", "GDP2019", "GDPpercap", "changedjob", 
#"lostjob1", "lostjob2", "unempoversamp", "lostjob", "logHHINCOME" )])






# ####################    LASSO


lasso.model = cv.glmnet(x, y, type.measure = "mse")

#plot it
plot(lasso.model)
#predicts lowest mse with 5 variable model

#checking coef's for lasso
coef(lasso.model)



# ################# RIDGE

ridge.model = cv.glmnet(x,y, type.measure='mse', alpha=0)
#plot it
plot(ridge.model)
#selects 8 model but with log-L of -3.something
coef(ridge.model)




# ################# PCR doesn't work with current x,y structure
pcr.fit <- pcr(y ~ x, data = cpshealth, scale = TRUE, validation = "CV")

#error: Error in if (any(abs(sdtrain) < .Machine$double.eps^0.5)) warning("Scaling with (near) zero standard deviation") : missing value where TRUE/FALSE needed




# ################### SVM

#factoring Y so that it can be used in SVM

dat = data.frame(x=x, y=as.factor(y))

svm.fit = svm(y~ ., data = dat, kernal = 'linear', cost = 10, scale = TRUE, na.action = na.omit)

#error: Error in if (any(co)) { : missing value where TRUE/FALSE needed



#OLS
# x = data.matrix(cpshealth[ c('logHHINCOME', 'SCHLCOLL', 'NFAMS', 'SEX', 'AGE', 'FAMSIZE', 'RACE', 'DISABWRK', 'rural')])


glm(y~ cpshealth$NFAMS + cpshealth$AGE + cpshealth$FAMSIZE + cpshealth$RACE +cpshealth$DISABWRK +cpshealth$rural + cpshealth$SCHLCOLL + cpshealth$SEX)

##not sure why but it won't let be use the log-income in the glm for OLS



















